---
title: "Survival Analysis of a Randomised Cancer Trial"
subtitle: "Does adjuvant chemotherapy delay recurrence and prolong survival in colon cancer?"
author: "Kingsley Amegah"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
    code_folding: hide
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.6, fig.align = "center")
suppressMessages({
  library(survival); library(survminer); library(dplyr); library(forcats)
  library(broom); library(gtsummary); library(ggplot2); library(randomForestSRC)
})
theme_set(theme_minimal(base_size = 12))
PAL <- c("#9ec7bd", "#e9b949", "#2a9d8f")   # Obs, Lev, Lev+5FU
ACC <- "#2a9d8f"
dat <- read.csv("data/colon_clean.csv", stringsAsFactors = TRUE)
dat$rx <- factor(dat$rx, levels = c("Obs", "Lev", "Lev+5FU"))
os  <- subset(dat, etype == 2)   # overall survival (death)
rfs <- subset(dat, etype == 1)   # recurrence-free survival
```

# Background and question

A central job of data science in health is to answer one question rigorously:
**does a treatment actually work, and for whom?** The cleanest evidence comes
from a randomised controlled trial (RCT), where patients are assigned to
treatments by chance so the groups are comparable.

This report analyses a real, published RCT: the **colon cancer adjuvant-therapy
trial** (Moertel and colleagues). After surgery, **929 patients** with stage B
or C colon cancer were randomly assigned to one of three arms:

- **Obs**, observation only (the control),
- **Lev**, levamisole,
- **Lev+5FU**, levamisole plus fluorouracil.

Two outcomes were followed over time: **recurrence** of the cancer and
**death**. The analytical challenge is that many patients had not had the event
by the end of follow-up (their times are *censored*), so ordinary averages do
not work. This is what **survival analysis** is built for.

# The data and randomisation check

Because treatment was randomised, the arms should look similar at baseline. A
balance table is the first thing a trial statistician checks.

```{r baseline}
os |>
  select(rx, age, sex, nodes, node4, differ, extent, obstruct, perfor) |>
  tbl_summary(by = rx,
              label = list(age ~ "Age (years)", nodes ~ "Positive nodes",
                           node4 ~ "Nodal burden", differ ~ "Differentiation",
                           extent ~ "Local extent", obstruct ~ "Obstruction",
                           perfor ~ "Perforation", sex ~ "Sex")) |>
  add_p() |>
  modify_header(label ~ "**Characteristic**") |>
  modify_caption("**Baseline characteristics by treatment arm**") |>
  as_kable()
```

The arms are well balanced: no baseline characteristic differs meaningfully
across them (all p-values are large). That is exactly what randomisation should
deliver, and it means any later difference in outcomes can be attributed to
treatment rather than to sicker patients landing in one arm.

# Kaplan-Meier survival curves

The Kaplan-Meier estimator turns censored follow-up times into a survival curve:
the share of patients still event-free as time passes. We compare arms with the
**log-rank test**.

## Overall survival (time to death)

```{r km-os}
fit_os <- survfit(Surv(time_yrs, status) ~ rx, data = os)
ggsurvplot(fit_os, data = os, palette = PAL, conf.int = TRUE,
           risk.table = TRUE, pval = TRUE, censor = FALSE,
           xlab = "Years from randomisation", ylab = "Overall survival",
           legend.title = "Arm", legend.labs = c("Obs", "Lev", "Lev+5FU"),
           risk.table.height = 0.28, ggtheme = theme_minimal(base_size = 12))
```

```{r km-os-stats}
lr_os <- survdiff(Surv(time_yrs, status) ~ rx, data = os)
p_os <- 1 - pchisq(lr_os$chisq, df = length(lr_os$n) - 1)
med <- summary(fit_os)$table[, "median"]
cat("Log-rank p-value (overall survival):", signif(p_os, 3), "\n")
cat("Median survival not reached where shown as NA (over half still alive).\n")
print(round(med, 2))
```

The **Lev+5FU** curve sits clearly above the other two, and the log-rank test
confirms the arms differ significantly. More than half of the Lev+5FU patients
were still alive at the end of follow-up, so their median survival is not even
reached. Levamisole alone tracks the control arm closely.

## Recurrence-free survival (time to recurrence)

```{r km-rfs}
fit_rfs <- survfit(Surv(time_yrs, status) ~ rx, data = rfs)
ggsurvplot(fit_rfs, data = rfs, palette = PAL, conf.int = TRUE,
           risk.table = TRUE, pval = TRUE, censor = FALSE,
           xlab = "Years from randomisation", ylab = "Recurrence-free survival",
           legend.title = "Arm", legend.labs = c("Obs", "Lev", "Lev+5FU"),
           risk.table.height = 0.28, ggtheme = theme_minimal(base_size = 12))
```

The same story appears earlier and even more strongly for recurrence: the
combination arm keeps far more patients cancer-free, and most of its benefit is
in preventing the cancer from coming back in the first place.

# Cox proportional-hazards model

The Kaplan-Meier curves compare arms on their own. The **Cox model** estimates
the effect of treatment while adjusting for prognostic factors, and reports a
**hazard ratio (HR)**: the relative rate of death. HR below 1 means lower risk.

```{r cox}
cox <- coxph(Surv(time_yrs, status) ~ rx + age + sex + nodes + extent +
               differ + obstruct + perfor, data = os)
tbl_regression(cox, exponentiate = TRUE,
               label = list(rx ~ "Treatment arm", age ~ "Age",
                            sex ~ "Sex", nodes ~ "Positive nodes",
                            extent ~ "Local extent", differ ~ "Differentiation",
                            obstruct ~ "Obstruction", perfor ~ "Perforation")) |>
  modify_caption("**Adjusted hazard ratios for death (Cox model)**") |>
  as_kable()
```

```{r cox-hr}
hr <- tidy(cox, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == "rxLev+5FU")
cat(sprintf("Lev+5FU vs Obs: HR = %.2f (95%% CI %.2f to %.2f), p = %.4f\n",
            hr$estimate, hr$conf.low, hr$conf.high, hr$p.value))
```

After adjusting for age, nodal burden, tumour extent and differentiation, the
combination arm reduces the rate of death by roughly **one third** compared with
observation, and the effect is statistically significant. The number of positive
lymph nodes and the local extent of the tumour are the strongest prognostic
drivers, as clinical knowledge would predict.

```{r forest, fig.height=6}
ggforest(cox, data = os, main = "Hazard ratios for death (95% CI)")
```

# Checking the proportional-hazards assumption

The Cox model assumes a treatment's effect on the hazard is roughly constant
over time. We test this with scaled Schoenfeld residuals.

```{r ph}
zph <- cox.zph(cox)
print(zph)
```

```{r ph-plot, fig.height=4.2}
plot(zph[1], col = ACC, lwd = 2)
abline(h = 0, lty = 2, col = "grey50")
title("Schoenfeld residuals for treatment arm")
```

The global test and the treatment term are not significant, so the
proportional-hazards assumption is reasonable for the treatment effect: the
hazard ratios above are a fair summary across the follow-up period.

# A machine-learning benchmark: Random Survival Forest

Cox is the clinical standard, but it assumes effects are linear and additive. A
**Random Survival Forest (RSF)** is a machine-learning model that captures
non-linearities and interactions automatically. We compare the two on
discrimination, measured by the concordance index (C-index, higher is better),
and read off which factors the forest finds most important.

```{r rsf}
set.seed(42)
rsf_dat <- os |>
  select(time_yrs, status, rx, age, sex, nodes, extent, differ,
         obstruct, perfor) |> na.omit()
rsf <- rfsrc(Surv(time_yrs, status) ~ ., data = rsf_dat,
             ntree = 600, importance = TRUE)
c_rsf <- 1 - rsf$err.rate[rsf$ntree]
c_cox <- summary(cox)$concordance[1]
cat(sprintf("C-index, Cox model           : %.3f\n", c_cox))
cat(sprintf("C-index, Random Survival Forest: %.3f\n", c_rsf))
```

```{r vimp, fig.height=4.6}
vi <- sort(rsf$importance, decreasing = TRUE)
vidf <- data.frame(var = names(vi), imp = as.numeric(vi))
ggplot(vidf, aes(reorder(var, imp), imp)) +
  geom_col(fill = ACC) + coord_flip() +
  labs(title = "Random Survival Forest: variable importance",
       x = NULL, y = "Permutation importance (higher = more predictive)")
```

The two models reach **very similar discrimination**, which is reassuring: the
straightforward Cox model is not leaving much on the table here. The forest
ranks the same drivers at the top (nodal burden, tumour extent, age), and it
also picks up treatment arm as informative. When a simple, interpretable model
matches a flexible one, the simple model is usually the right choice to report.

# Interpretation and clinical recommendation

```{r summary, echo=FALSE}
risk_red <- (1 - hr$estimate) * 100
```

- **The combination works.** Adjuvant **levamisole plus fluorouracil** clearly
  delays recurrence and prolongs survival in stage B/C colon cancer, cutting the
  rate of death by about **`r round(risk_red)`%** versus observation after
  adjustment, consistent across the Kaplan-Meier curves, the log-rank tests and
  the Cox model.
- **Levamisole alone does not.** Its survival curve sits on top of the control
  arm; the active ingredient is the fluorouracil combination.
- **Most of the benefit is in preventing recurrence**, which is where the curves
  separate earliest and widest.
- **Target the highest-risk patients.** Patients with more than four positive
  nodes and greater local extent have the worst prognosis and the most to gain,
  so they are the clearest candidates for the combination regimen.

# Limitations

This is a single historical trial, so the specific drug regimen reflects the
standard of its era rather than today's protocols; the value here is the method,
not a current treatment recommendation. Some covariates have missing values
(handled by complete-case analysis for the models). Survival analysis estimates
associations adjusted for measured factors only; unmeasured confounding is
controlled by randomisation for the treatment comparison but not for the
observational prognostic factors.

---

*Built in R (survival, survminer, randomForestSRC, gtsummary) from the colon
cancer adjuvant-therapy trial, a real published RCT with openly available
patient-level data. Author: Kingsley Amegah.*
